---
title: "How to write functions using dplyr"
author: "Naas van Heerden"
date: "10/11/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### The Why

Writing functions with dplyr makes a lot of sense because you can automate the living daylight out of most data wrangling tasks! You also get to use a familiar syntax within your function body (if you are a regular dplyr user). 

#### Here is an example:

Let's load the necessary packages: 

```{r message=FALSE}
library(dplyr)
library(rlang)
```

First we create a data set.

```{r}
x <- data.frame(a=c(1:10), b=c(10:1), c=c(1:10))
x
```

The  **objective** of the task we want to perform with this dataset is as follows: Imagine `a` is some kind of index or time variable. We would like to sum all the other variables in the data set (we don't necessarily know how many there are) to a total and then calculate the percentage of each variable makes up at each point of `a` of that total. 

So it makes sense for us assign the number of columns to a variable, as we want to automate the task. 

```{r}
ncol <- ncol(x)
ncol
```

Now we add the total variable to the data frame using the ncol object (automation in mind)

```{r}
x$total <- apply(x[,c(2:ncol)], 1, sum)
x
```

To add the variables that represent the percentages of the total is quite easy: 
```{r}
mutate(x, perc_b = b / total) %>%
  mutate(perc_c = c / total)
```

But what if we want to write a function or script to do this and we don't know how many variables there are in the data frame that is given to fed into the function or script? 

What we want to do is essentially: 

for all the variables in the data set except `a` and `total` do this: 
`mutate(perc_variablename = variablename / total)`

So let's try it for one variable only first: 

```{r error=TRUE}
variablename <- "b"
mutate(paste0("perc_", variablename) = variablename / total)
```

This gives us an error message. I'm not going to explain the theory behind it. You can read more about it [here](https://cran.r-project.org/web/packages/dplyr/vignettes/programming.html). 

This is how you fix it. Note that the LHS of the mutate expression needs to evaluate to a string. 

```{r}
variablename <- "b"
# total is a variable in the data frame, x, so it needs to be quoted.
total <- quo(total)

# the new variable name (LHS) should evaluate to a string:
new_name <- paste0("perc_", variablename)

# when using the variable name on the RHS it should evaulate to a variable name
# :)
new_variablename <- as.name(variablename)

mutate(x, !!new_name := UQ(new_variablename) / UQ(total))
```

Now to automatically do this action for how many variables there are, we do the following

```{r}
for (i in 2:ncol) {

name <- as.name(names(x[i]))
nname <- paste0("perc_", names(x[i]))

total <- quo(total)

x <- mutate(x, !!nname := UQ(name) / UQ(total))
}

x
```

So we could easily put the above into a function.  

```{r}
add_perc <- function(data) {
  ncol <- ncol(data) - 1
  for (i in 2:ncol) {

    name <- as.name(names(data[i]))
    nname <- paste0("perc_", names(data[i]))

    total <- quo(total)

    data <- mutate(data, !!nname := UQ(name) / UQ(total))
  }
  return(data)
}

# recreate x 
x <- data.frame(a=c(1:10), b=c(10:1), c=c(1:10))
x$total <- apply(x[,c(2:ncol)], 1, sum)

add_perc(x)
```

